Keep your Laziness in Check

### 2018-01-23 paper/main.tex
rewrote intro again

\hzh{Rewrite counter: 1}
\hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
\hzh{We can write a hand-crafted spec for take, can we show that we can create this spec by composing the spec for foldr (but not using the one with foldl)}
\hzh{implement take in 3 ways: foldl, foldr and explicitly as the opening example, and then ask if they're equivalent? (they're not!)}
\hzh{Okasaki has 2 functions that appends a list to the reversal of another list, one of which is too strict and the other one is just the right amount of lazy (thus incremental)}
\hzh{The amortized one is hard to specify exactly, but we can introduce approximate specification and specify it approximately}
\hzh{we need to carefully avoid saying this is a tool for specifying the time complexity of functions}
\hzh{Implementing functional programs that make clever use of laziness is very easy to get wrong, and we provide a solution to make sure the programmer got it right. We almost have no way of testing it right now before StrictCheck.}
\hzh{Email Ryan Trinkle and Dan Winograd-Cort about FRP and strictness, and try to use their examples as another problem we can solve in the intro}
\hzh{Lazy maps will mess up memoization if you're overly strict, use knapsack as an example}
\leo{I want to keep a counter on how many times we'll rewrite this intro :)}

John Hughes has argued adequately that lazy evaluation brings tremendous power
to modularization of functional programming in his seminal work \textit{Why
Functional Programming Matters} ~\cite{WhyFPMatters}. Haskell is one of the most
popular programming languages that gives its programmers the power of lazy
evaluation.
%
Unfortunately, reasoning about the performance of lazy programs in Haskell is
difficult. Programmers often accidentally make a program too lazy or too strict,
and as a result, their programs behave differently than expected.

@@

For example, consider the following simple Haskell implementation of the \lk{sum} function:
\leo{I created an environment for this so that we can tweak them all as needed}
\begin{inlinecode}
sum = foldl (+) 0
\end{inlinecode}
%
Although this seemingly innocent implementation of sum will produce correct
results on small lists of numbers, it will quickly run out of memory on larger
lists. The culprit here is the laziness behavior of \lk{foldl}---it does not
force the accumulated result during the fold. Instead, calling this \lk{sum} on
a list produces a thunk that is linear in the size of of the input list, and
this thunk will not be evaluated until the result of sum is used in another
computation.
%
We can fix this space leak for \lk{sum} using a strict fold, \lk{foldl'}, which
eagerly forces the accumulator, and avoids creating a space-leaking thunk.

However, using \lk{foldl'} is not sufficient to avoid space leaks in
general. Consider a single-traversal computation of an average where we keep
both a running sum and a running counter in an accumulator.
%
\hzh{TODO: rewrite this, let bind the foldl}
\begin{inlinecode}
avg = uncurry (/) . foldl (\(s,n) x -> (s + x, n + 1)) (0,0)
\end{inlinecode}
%
In this case, both \lk{foldl} and \lk{foldl'} construct thunks linear in the
size of the input list, since forcing a lazy pair only evaluates it to its weak
head normal form, and it does not evaluate either component of the pair.

How, then, can we test and catch these laziness bugs before they
manifest themselves in production? Is it possible to precisely specify
laziness behavior just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in Haskell, and
propose a system of creating specifications of lazy Haskell programs. We
implement these ideas as a Haskell library.

>>>>

\hzh{Rewrite counter: 2}
\hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
\hzh{We can write a hand-crafted spec for take, can we show that we can create this spec by composing the spec for foldr (but not using the one with foldl)}
\hzh{implement take in 3 ways: foldl, foldr and explicitly as the opening example, and then ask if they're equivalent? (they're not!)}
\hzh{Okasaki has 2 functions that appends a list to the reversal of another list, one of which is too strict and the other one is just the right amount of lazy (thus incremental)}
\hzh{The amortized one is hard to specify exactly, but we can introduce approximate specification and specify it approximately}
\hzh{we need to carefully avoid saying this is a tool for specifying the time complexity of functions}
\hzh{Implementing functional programs that make clever use of laziness is very easy to get wrong, and we provide a solution to make sure the programmer got it right. We almost have no way of testing it right now before StrictCheck.}
\hzh{Email Ryan Trinkle and Dan Winograd-Cort about FRP and strictness, and try to use their examples as another problem we can solve in the intro}
\hzh{Maps will mess up memoization if you're overly strict, use knapsack as an example}
\leo{I want to keep a counter on how many times we'll rewrite this intro :)}

Lazy evaluation brings tremendous power to modularization of functional
programming ~\cite{WhyFPMatters}. Haskell is one of the most popular programming
languages that gives its programmers the power of lazy evaluation.

To demonstrate the benefits of lazy evaluaation, let us consider a dynamic
programming solution of the knapsack problem in Haskell.
%
The inputs of the knapsack problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the weight and the
value of the $i$th item. The solution presents the maximum value achievable
using a subset of the items such that their combined weight is under a limit
$W$.

The typical dynamic programming approach for knapsack maintains a table $OPT[][]$,
where the entry $OPT[j][k]$ keeps the best value achievable using only items
$1\dots j$ and a weight limit of $k$. The table satisfies the following
recurrence relation: $$OPT[j][k] = \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

With typical imperative languages, we look for shared sub-problems of the
original knapsack problem, and then we design a path along the solution table
that maximizes sharing while filling out the answers.
%
However, lazy evaluation automatically finds this path if we
use \lk{Data.Map.Lazy} and setup the solution table with the recurrence
relation \hzh{We can probably make this code prettier}
\begin{inlinecode}
ws, vs :: [Int]
n,  w  :: Int

solution :: Map (Int, Int) Int
solution = fromList [ ((j, k),
                       if (j-1) < 0 || k - ws !! j < 0
                       then 0
                       else max (solution ! (j-1, k))
                                (solution ! (j-1, k - ws !! j) + vs !! j)
                      )
                    | j <- [0..2], k <- [0..50] ]
\end{inlinecode}

Using the lazy variant of \lk{Map} here is critical for the correctness
of \lk{solution}. If we used \lk{Data.Map.Strict} instead, then \lk{fromList}
would have eagerly evaluated the \lk{if} expression. The \lk{if} expression
contains value-level recursion to itself, and this would manifest as an infinite
loop at runtime. This is an incredibly subtle bug related to laziness.

Unfortunately, reasoning about laziness in Haskell is difficult. Programmers
often accidentally make a program too lazy or too strict, and as a result, their
programs behave differently than expected.

@@

How, then, can we test and catch these laziness bugs during development? Is it
possible to precisely specify laziness behavior just as we would for functional
correctness?
%
In this paper, we demonstrate techniques for observing laziness in Haskell, and
propose a system of creating specifications of lazy Haskell programs. We
implement these ideas as a Haskell library.
